%%%%%%%%%%%%%%%%  NUMERICAL RESULTS %%%%%%%%%%%%%%%%%%%%%
\chapter{Numerical Results} \thispagestyle{chapterpage}
\label{chapter:numerical_results}

\todoinline{Run tests with lower tolerance. Might favor Newton/JTR}

Reservoir simulation packages are large and complex programs since input of well specifications, grid parameters, fluid definitions, etc., must be supported by the code base, along with all kinds of utility functions. In order to test the numerical methods described in Chapter \ref{chapter:numerical_methods} we use the open source simulator supplied by the Open Porous Media initiative, or \opm, see \citet{opm_2014}. Section \ref{section:opm_package} starts this chapter with a brief overview of the \opm transport solver classes used as a basis for the new root finder implementations (see Section \ref{section:root_finders}). Section \ref{section:test_cases} presents a number of tests cases and the corresponding numerical results from running the cases with the root finders from Section \ref{section:root_finders}.

%%%%%%%%%%%%%%%%  THE OPM PACKAGE %%%%%%%%%%%%%%%%%%%%%
\section{The OPM Package}
\label{section:opm_package}
\todoinline{Fix the over/underfull lines}
The \opm package provides a range of modules for grid handling, polymer injection, upscaling methods, and more. The \code{opm-core} module contains basic grid and well handling, and IO utilities, along with pressure and transport solvers for solving the porous media fluid flow problems described in Section \ref{section:fluid_models}. In fact, the \opm package implements the exact numerical methods described in Section \ref{section:seq_splitting_method} through the \code{Opm::IncompTpfa} pressure solver and the \code{Opm::TransportSolverTwophaseReorder} transport solver classes using the Regula Falsi, see Section \ref{section:regula_falsi}, for the single cell problems resulting from  reordering as described in Section \ref{section:reordering}. The class \code{Opm::TransportSolverTwophaseReorder} implements the functionality needed to solve the residual equations in Equation (\ref{eq:residual_two_phase_transport}) and (\ref{eq:residual_two_phase_transport_gravity}). The class is instantiated with the static properties of the simulation, such as a grid specification and a fluid model. At each iteration of the sequential splitting method, as outlined in Algorithm \ref{algorithm:sequential_splitting}, a new saturation field is computed by calling the method \code{solve(\dots)} on a persistent \code{Opm::TransportSolverTwophaseReorder} object. The arguments to the solve method includes the time step $\Delta t$, and an instance of the class \code{TwophaseState} containing the saturation, flux, and other state information for every cell $V \in \mathcal{T}$. The \code{solve} method proceeds to computes the ordering of the flux graph based on the flux values obtained by solving the pressure equation. This ordering leads to a set of pseudo cells consisting of one or more regular cells, as described in Section \ref{section:reordering}. The new saturation field is finally obtained by iterating over all the pseudo cells and solving each subproblem by either the single cell root finders or the multi cell solver depending on the number of grid cells in each pseudo cell. The single cell root finders are the main focus of this work. The transport equation residuals are implemented by two structs, Equation (\ref{eq:residual_two_phase_transport}) in the struct \code{Residual} and Equation (\ref{eq:residual_two_phase_transport_gravity}) in the struct \code{GravityResidual}. These structs are supplied with an ()-operator taking the update saturation $S_V^{n+1}$ as an argument and returning the residual value. The following section walks through a simple reservoir simulation setup using the \opm library.
\input{spe10_perm}

%%%%%%%%%%%%  TEST CASES  %%%%%%%%%%
\section{Test Cases}
\label{section:test_cases}
%%%%%%%%%%%%  TEST PROCEDURE  %%%%%%%%%%
\subsection{Test Procedure}
\label{section:test_procedure}
In order to test the efficiency of the single cell solvers the \opm library was installed on a server with Intel\textregistered{}  Xeon\textregistered{}  X7542 CPUs running at \unit[2.67]{GHz} with a \unit[18432]{kB} cache size. The server has \unit[252]{GB} of available ram. All 2D test have been run using the C++ driver program included in Listing \ref{listing:test_driver_2D} Appendix \ref{appendix:test_driver}. For the 3D tests the code in Listing \ref{listing:test_driver_3D} was used. Further, all test cases are are checked against the reference Regula Falsi solver, see Section \ref{section:regula_falsi}, to ensure that the same solution is found. The iteration count for each solver is also recorded along with the solution updates to determine the  convergence speed versus iteration for each method. Finally, the CPU time averaged over the number of cells in the grid is reported to check the overall performance of each root finder. In the following we will test these methods for solving the single cell residual: Brent (B), Regula Falsi (RF), Ridders (R), Approximate Jenny Trust Region (TR*),  Globalized Newton (GN), and Jenny Trust Region (TR). Note the method name abbreviations in the parentheses.
\input{sat_hist_hom}

%%%%%%%%%%%%  QUARTER FIVE SPOT %%%%%%%%%%
\subsection{Case A: Quarter Five Spot}
\label{section:caseA}
\todoinline{Error plot against reference solver for hom. Q5}
The \emph{quarter five spot}, abbreviated Q5, is a quadratic 2D domain with a source in one corner and a sink in the opposite corner along the diagonal. Here the domain has dimensions $\unit[120]{m}\times\unit[120]{m}\times\unit[10]{m}$. The fluid density is set to $\unitfrac[1000]{kg}{m^3}$, porosity $\unit[0.5]{}$ and the formation has homogeneous permeability $\unit[10]{mD}$. The simulation is run on a $20\times 20$ grid with a strong source in one corner cell and a sink in the diagonally opposite corner, both of magnitude $\unitfrac[180]{m^3}{s}$. The viscosity is varied through three cases; $\mu_w = \unit[1]{cP}$ and $\mu_o= \unit[1]{cP}$, $\mu_w = \unit[1]{cP}$ and $\mu_o= \unit[10]{cP}$, and $\mu_w = \unit[10]{cP}$ and $\mu_o= \unit[1]{cP}$. Figure \ref{fig:sat_hist_hom} shows an example of an advancing saturation profile from a simulation of the homogeneous Q5 problem. Because of the uniform permeability and the quadratic geometry the saturation profile is symmetric around the diagonal between the source and sink.

%%%%% 120 by 120 %%%%%%%
\subsubsection{120 m by 120 m comments}
The total iteration count spent when solving the Q5 problem is presented in Figures \ref{fig:caseA_iterations_t_5} and \ref{fig:caseA_iterations_t_5} for all root finders. We include the viscosity ratio because of the significant influence it has on the shape of $f_w$, as indicated by Figure \ref{fig:fractional_flow_wrt_viscosity_ratio}. The trend in the data is that Brent's method uses the highest number of iterations, while the trust region scheme needs significantly fewer iterations than all other methods. The data also indicates that the average iteration count is lower for larger $M$-values, and that the differences between the methods are smaller.

The total CPU running times are shown as a function of the time step $\Delta t$ for three different viscosity ratios $M$ in Figure \ref{fig:q5_homperm_10}. Again we include different $M$ values. With $M = 0.1$ all methods have similar performance up to around $\Delta t = \unit[20]{d}$, as shown in Figure \ref{fig:q5_cputime_perm_h_10_m_1_10_i_4}. At this point the regula falsi method, and a little later also the approximate trust region scheme, starts to spend more CPU time than the other methods. This trend is consistent over all subsequent time step sizes. We note that the other methods has equal performance for all time steps.
Next, Figure \ref{fig:q5_cputime_perm_h_10_m_1_1_i_4} shows the results for $M = 1$. Here all the root finders uses equal CPU time over the whole range of time steps.
Finally, $M = 10$ gives the results in Figure \ref{fig:q5_cputime_perm_h_10_m_10_1_i_4}. In this case the regula falsi is faster than all the other methods for $\Delta t \gtrapprox \unit[20]{d}$. Further we observe that the Ridders and Brent methods break of last at $\Delta t \approx \unit[20]{d}$, while the trust region scheme breaks of at $\Delta t \approx \unit[5]{d}$, and the approximate trust region scheme does the same at $\Delta t \approx \unit[10]{d}$.

\input{q5-cputvsdt-s-x-T-300-m-x-dim-120-120-20-20-perm-h-10-i-180}
\input{q5-iterations-s-x-T-300-m-x-dim-120-120-20-20-perm-h-10-i-180}

\subsubsubsection{Discussion}
A comparison of the iteration count and total CPU time results for test case A, shown in Figure \ref{fig:caseA_iterations_t_5} and \ref{fig:q5_cputime_perm_h_10_i_4}, respectively, highlights the differences in computational complexity for the different root finders. That is, even with the quite significant variation in total amount of iterations the CPU time is comparable for all root finders over the range of tested parameters. For instance, the trust region methods, which are essentially Newton methods, converge fast in terms of number of iterations due to the quadratic convergence of the Newton-Raphson scheme. But, since each iteration requires two function evaluations, one for the function itself and one for the derivative, the gains in iterations are balanced by a high overhead in each iteration. Similarly, the Brent method has ``only'' superlinear convergence and thus uses a very high number of iterations, but since it requires just one function evaluation per iteration the total CPU time spent is again normalized. Similar considerations can be used to explain the iteration count versus CPU time results for the other root finders.

\begin{itemize}
\item NB: The following observations are from simulation with $\Delta t = \unit[50]{d}$, 120 by 120 m!
\item $M = 0.1$: $\mu_{e_r} = 0.0814, \sigma_{e_r} = 0.0677$, $\mu_{S_r} = 0.3549, \sigma_{S_r} = 0.1832$
\item $M = 1$: $\mu_{e_r} = 0.1219, \sigma_{e_r} = 0.1242$, $\mu_{S_r} = 0.4815, \sigma_{S_r} = 0.3104$
\item $M = 10$: $\mu_{e_r} = 0.1474, \sigma_{e_r} = 0.2192$, $\mu_{S_r} = 0.5224, \sigma_{S_r} = 0.4210$
\item Large $M$ gives small $f_w$, i.e. lower flow of water out of cell. This gives higher water saturation on average.
\end{itemize}

Figure \ref{fig:q5_root_saturations_t_50} shows the distribution of the cell saturations $S_V^{n+1}$ for all time steps $\Delta t$ and three different viscosity ratios. The mean and standard deviation of the distributions, shown in Table \ref{table:q5_saturations_statisitcs_t_50}, indicates that the viscosity ratio $M$ has significant impact on the final saturation distribution in the domain. It seems that a large $M$ gives a larger average water saturation in the cells, while smaller $M$ gives smaller average saturation values. The trend in the shape of the distribution is even clearer, with larger values of $M$ pushing the distribution to the right and leaving a larger fraction of saturations near zero. These observations are readily explained by again noting that small $M$ gives values of the fractional water flow function $f_w$ closer to one, while larger $M$ keeps $f_w$ close to zero, as shown in Figure \ref{fig:fractional_flow_wrt_viscosity_ratio}. The $f_w$ measures the water flow, and scales the outgoing flux in the transport residual, Equation (\ref{eq:residual_two_phase_transport_simple}). Thus, large $f_w$ values gives a large flow of water out of the cell, while small gives a small flow. A large outflow of water will lead to smaller saturation values in the cell, and vice versa, which is exactly what we observe in Figure \ref{fig:q5_root_saturations_t_50}.

\input{q5_saturations_statisitcs_t_50_i_180}

\textbf{The results from the saturation statistics indicates that the initial guess errors are significant for large $M$, i.e. $f_w$ small. Small outflow gives large saturation values, i.e. large M gives large saturation values. A sharp front, since the flow enters the cell easily but leaves slowly. Causes a large percentage of zero cells. Small M gives more equal flow in and out of the cell, causing a smeared front and more uniform saturation values. NB: Does small M imply flux values of comparable size in and out of the cell?  How does this correlate with the discussion from the convergence tests?}

\input{q5_root_saturations_i_180}
\input{q5_initial_guess_errors_i_180}

\clearpage
%%%%% 10 by 10 %%%%%%%
\subsubsection{10 m by 10 m comments}
The total number of iterations spent by each root finder when solving the Q5 problem is shown as a function of $\Delta t$ and the viscosity ratio $M$ in Figure \ref{fig:q5_iterations_perm_h_10_i_4}. The Brent methods spends the most iterations, while the regula falsi is a close second. Ridders method needs more iterations than the two trust region schemes. For $\Delta t \lessapprox 60$ the precise trust region scheme is the most efficient with the approximate trust region scheme a close second, while $\Delta t \gtrapprox 60$ favours the approximate trust region scheme. Qualitatively these observations are consistent for all tested viscosity ratios. Note however that the quantitative difference between the number of iterations spent by the various root finders is influenced by $M$. The qualitative difference between the methods are about as expected based on their theoretical convergence rates. The trust region schemes benefit from the quadratic local convergence of the underlying Newton scheme, while the other methods with only superlinear convergence need more iterations to obtain the desired precision. The advantage of the trust region schemes over the other methods seem quite significant.

\input{q5-iterations-s-x-T-300-m-x-dim-10-10-20-20-perm-h-10-i-4}

The total CPU time spent when solving the Q5 problem is shown in Figure \ref{fig:q5_cputime_perm_h_10_i_4} for all root finders and different viscosity ratios. Starting with Figure \ref{fig:q5_cputime_perm_h_10_m_1_10_i_4} for $M = 0.1$ we see that the CPU times for all methods are practically indistinguishable up to $\Delta t \approx \unit[20]{d}$. At this point the precise trust region method becomes faster than the rest. This holds until $\Delta t \approx \unit[50]{d}$ where the regula falsi method also lowers its CPU time. Finally, the Brent method speeds up somewhere between $\Delta t = \unit[50]{d}$ and $\unit[50]{d}$. The Ridders method and the approximate trust region method have similar, and the lowest, performance for all time steps. The exception is the very last time step where the approximate trust region scheme jumps to the level of the fastest algorithms.

The timing results with $M = 1$ are shown in Figure \ref{fig:q5_cputime_perm_h_10_m_1_1_i_4}. The results in this case are less clear cut than with $M = 0.1$, with run times jumping up and down between time steps. We note a few trends though; First, the Ridders and approximate trust region methods have fairly consistent low timing results. Second, the regula falsi and precise trust region schemes perform worse than the two aforementioned methods for the mid range of time step values, between $\Delta t \approx \unit[20]{d}$ and $\Delta t \approx \unit[100]{d}$. After that they have good performance.

Finally, Figure \ref{fig:q5_cputime_perm_h_10_m_10_1_i_4} shows the results with viscosity ratio $M = 10$. Here the precise trust region method is consistently the most efficient method after all other methods break away between $\Delta t = \unit[20]{d}$ and $\unit[20]{d}$. The other methods have similar performance for the rest of the tested time steps.

A general observation is that all methods seems to have similar performance for the smallest time steps, up until around $\Delta t \approx \unit[20]{d}$.

\begin{itemize}
\item General impression from watching $M=0.1$ residuals: Root often close to inflection point.
\item -> Maybe check number of inflection point crossings?
\item $M=0.1$: Last time step has lots of relatively linear residuals
\item -> Perhaps trust region is comparably slow on linear residuals?
\end{itemize}

\input{q5-cputvsdt-s-x-T-300-m-x-dim-10-10-20-20-perm-h-10-i-4}

\input{q5_saturations_statisitcs_t_50_i_4}
\input{q5_root_saturations_i_4}
\input{q5_initial_guess_errors_i_4}

\clearpage
%%%%%%%%%%%%%%  TARBERT 2D  %%%%%%%%%%%%%%%%%%%
\subsection{Case B: Tarbert 2D}
\label{section:caseB}
\todoinline{Comment the results from Tarbert 2D}
The Q5 problem is used again but with a more realistic inhomogeneous permeability distribution taken from the second SPE10 data set \citep{spe10_2000}. The residuals resulting from such permeability distributions typically have different characteristics than the ``homogeneous residuals'' some of which might favor different root finders. The second SPE10 data set consists of scalar permeabilities in the $x$-, $y$- and $z$-directions on a three dimensional grid with $60\times 220\times 85$ cells, with the top 35 layers being part of the Tarbert formation and the bottom 50 the Upper Ness formation.The fine scale permeability grid cells are $\unit[20]{ft}\times \unit[10]{ft}\times \unit[2]{ft}$ in size. Figure \ref{fig:spe10_perm} shows the logarithm of the x-direction permeabilities for the entire domain, from which the $\unit[120]{m}\times \unit[120]{m}$ region starting at $(x,y) = (0,0)$ in the first layer of the Tarbert formation is chosen for the numerical tests. Since this case has $220\times 60 = 13200$ cells versus the $400$ cells in case A we expect the iteration and CPU times to be significantly larger for case B.

\input{sat_perm_tarbert_2D}

%%%%% 60 by 60 %%%%%%%
\subsubsection{60 m by 60 m comments}

\input{spe10-cputvsdt-s-x-T-30-m-x-dim-60-60-10-60-220-1-perm-i-0-0-0-i-50}

\clearpage
%%%%% 10 by 10 %%%%%%%
\subsubsection{10 m by 10 m comments}

Figure \ref{fig:spe10_iterations_perm_i_0_0_0_i_50} shows the total iteration count used when solving case B using the different root finders and for varying $M$. As noted, the iteration count is orders of magnitude larger than for case A with a range from around $9\times 10^4$ up to around $1.5\times 10^7$ iterations. Again the lowest iteration count is obtained by the trust region schemes, with an advantage to the precise trust region method. The other methods also follow the pattern observed in case A, with the Brent method using the most iterations, and the regula falsi and Ridders methods slightly lower. We note that for $\Delta t \lessapprox \unit[30]{d}$ the regula falsi method has a slight advantage, while the Ridders method is better for larger time steps. 

\input{spe10-iterations-s-x-T-400-m-x-dim-10-10-60-220-perm-i-0-0-0-i-50}
%\input{spe10-cputvsdt-s-x-T-400-m-x-dim-10-10-60-220-perm-i-0-0-0-i-50}

\clearpage
%%%%%%%%%%%%%%%%  UPPER NESS 2D %%%%%%%%%%%%%%%%%%%
\subsection{Case C: Upper Ness 2D}
\label{section:caseC}
\todoinline{Comment the results from Upper Ness 2D}
Figure \ref{fig:spe10_perm} indicates that the Upper Ness formation has more severe local permeability variations compared to the Tarbert formation. These large local variations appear as \todo{show the effects of large local perm. gradients on the residual.}. Again we want to investigate how the various root finders perform on this special case, choosing the $\unit[360]{m}\times \unit[180]{m}$ region starting at $(x,y) = (0,0)$ in layer 45 of the SPE10 data set for the permeabilities.

\input{sat_perm_upperness_2D}

%%%%% 60 by 60 %%%%%%%
\subsubsection{60 m by 60 m comments}

\input{spe10-cputvsdt-s-x-T-30-m-x-dim-60-60-10-60-220-1-perm-i-36-0-0-i-50}
\input{spe10-cputvsdt-s-x-T-30-m-x-dim-60-60-10-60-220-1-perm-i-36-0-0-i-50}

\clearpage
%%%%% 10 by 10 %%%%%%%
\subsubsection{10 m by 10 m comments}

\input{spe10-iterations-s-x-T-400-m-x-dim-10-10-60-220-perm-i-36-0-0-i-50}
%\input{spe10-cputvsdt-s-x-T-400-m-x-dim-10-10-60-220-perm-i-36-0-0-i-50}

\clearpage
%%%%%%%%%%%%%%%% TARBERT - 3D %%%%%%%%%%%%%%%%%%%
\subsection{Case D: Tarbert 3D}
\label{section:caseD}

\todoinline{Permeability/saturation plot for Tarbert 3D}
\todoinline{Run Tarbert 3D with gravity - cputime}
\todoinline{Run Tarbert 3D with gravity - iterations}

\clearpage
%%%%%%%%%%%%%%%%  UPPER NESS - 3D %%%%%%%%%%%%%%%%%%%
\subsection{Case E: Upper Ness 3D}
\label{section:caseE}

\todoinline{Permeability/saturation plot for Upper Ness 3D}
\todoinline{Run Upper Ness 3D with gravity - cputime}
\todoinline{Run Upper Ness 3D with gravity - iterations}

\clearpage
%%%%%%%%%%%%%%%% CONVERGENCE TESTS %%%%%%%%%%%%%%%%%%
\section{Convergence Tests}
\label{section:numerical_results_convergence_tests}
The efficiency of the numerical methods in terms of convergence speed can highlight the properties of the numerical procedures. We again \todo{Comment briefly on the properties of the viscosity residual when it is introduced.} observe that the form of the viscosity dominated residual from Equation (\ref{eq:residual_two_phase_transport}) is determined by five parameters; the initial saturation in the cell, $S_V^{n}$, the time step to pore volume ratio $\tau = \frac{\Delta t}{m(V)\phi_V}$, the flux out of the cell $q_o$, the flux into the cell $q_i$, and finally the viscosity ratio $M$ as defined in Equation (\ref{eq:viscosity_ratio}), held constant at $M = 1$. Note that the cell saturation from the previous time step is used as initial guess for the root finders. Figures \ref{fig:conv_res_in_0_05_out_0_05} through \ref{fig:conv_res_in_0_5_out_0_8} shows a number of different convergence and residual plots obtained by varying these parameters. 

Figure \ref{fig:conv_res_in_0_05_out_0_05} is obtained with incoming flux at $\unitfrac[0.05]{m^3}{s}$ and outgoing flux at $\unitfrac[0.05]{m^3}{s}$. Note that $\tau = \unitfrac[6]{s}{m^3}$. Under the circumstances the residuals are fairly linear, and the initial guess $S_0$ is close to the root. The number of iterations for all tested root finders decreases when $S_0$ is increased. We also note that the Newton-like methods converge faster than the other methods. This plot indicates that for small flux values the initial guess strongly influences the residual bringing the root close to $S_0$. with a stronger effect for larger $S_0$. 

Setting the incoming flux to $\unitfrac[0.35]{m^3}{s}$ and the outgoing flux to $\unitfrac[0.35]{m^3}{s}$ we obtain Figure \ref{fig:conv_res_in_0_35_out_0_35}. The residual plots shown a stronger non-linear influence, a fact reflected in the increased iteration count. $S_0$ still influences the residual, but to a lesser extent. Again we observe that large initial guesses generally leads to a lower iteration count. The Newton-like method are, together with Ridders, the best performers.

Moving on the even larger flux values, we set the incoming flux to $\unitfrac[0.5]{m^3}{s}$ and the outgoing flux to $\unitfrac[0.8]{m^3}{s}$. Figure \ref{fig:conv_res_in_0_5_out_0_8} shows the resulting plots. The trend from the previous plots continues, in that the initial guess has less influence on the performance of the root finders, on average. Interesting exceptions are the Newton-like methods. They perform significantly better in terms of iteration count when the initial guess is close, in contrast with the other methods. We also note that the initial has very little influence on the position of the root.

\input{convergence2-s-x-M-1_000000-dtpv-6_000000-in--0_050000-out-0_050000-s0-x}
\input{convergence2-s-x-M-1_000000-dtpv-6_000000-in--0_350000-out-0_350000-s0-x}
\input{convergence2-s-x-M-1_000000-dtpv-6_000000-in--0_500000-out-0_800000-s0-x}
